Modeling Intervention Strategies for United States TB Control

### 2014-08-03 in_progress/presentations/paper/draft.tex
Updated draft

The majority of TB infections are caused by Mycobacterium tuberculosis, 
an airborne pathogen which primarily affects the lungs and pulmonary system.  
Individuals infected with this bacteria first develop a latent form of TB
referred to as latent tuberculosis infection (LTBI).  These individuals are
asymptomatic and not contagious, and thus, this infection is difficult to detect
and can persist for many years. The latent tuberculosis infection may progress
to active tuberculosis infection (active TB), at which point the patient becomes
symptomatic and contagious. This progression is most likely within the first two
years after contraction, but can happen throughout the rest of the lifespan of
the individual.  Left untreated, individuals with LTBI have a lifelong
probability of approximately 10\% to develop active TB.
\TODO{in total? Or after they've progressed through the first two years? We need
a new source for this, I think. It's not in the logical place of the WHO report.
This also obviously needs to be cited. }
The health risks for individuals with active TB are severe, and without medical
interventions, the 10 year mortality rate for the disease is between 20\% and
70\%, depending on the severity of the case and definition of a positive
diagnosis.
\cite{_who_2013}.

@@

The model was also extended to provide similarly sourced US health care system
(HCS) TB costs. To do so, we made the following treatment cost assumptions. A
single case of active TB was assumed to yield a \$14,000 US HCS cost, charged
immediately upon disease contraction. This cost is the weighted mean of the
costs of cases requiring hospitalization (49\%) and cases not requiring
hospitalization (51\%).  These cost estimates were found in \TODO{<+DYLAN COST
STUDY CITATION+>.} LTBI treatment costs were assumed to be \$468.00, charged
immediately upon successful treatment. These estimates were calculated based on
the cost of a successful treatment, and the typical adherence and efficacy of
said treatment \TODO{<+DYLAN COST STUDY CITATION, ADHERENCE CITATION+>.} Note
that active TB treatment costs are charged immediately upon disease contraction,
whereas LTBI treatment costs are charged only upon successful treatment. The
motivation for this distinction is that active TB treatment is mandated by the
US for all known cases and, further, most treatment costs are incurred at the
beginning of the treatment cycle. As such, charging for every known case,
immediately at contraction is a well motivated simplification. On the other
hand, latent TB is only rarely treated in discovered cases and the costs are
more evenly distributed. Furthermore, the choice to charge only upon successful
treatment yields a more conservative estimate of the success of any intervention
analyzed, as it will always underestimate the total US HCS TB cost burden.
Finally, it is worth noting that these estimates of US HCS treatment cost are
not intended to be fixed parameters. The code for this model is freely
available, and as new research amends the necessary costs and new methods of
treatment are developed, these values can be modified and the model re-run to
produce a revised prediction of the US HCS TB cost. In addition to US HCS cost,
the system was implemented so as to also track projected intervention
implementation cost, given user-inputted parameters relating to various possible
intervention cost strategies.  Cured cases of LTBI entering, TB deaths, total TB
cases, and total US HCS cost were also tracked discounted at 3\% annually for
cost benefit analysis. This discounting was converted to a continuous
differential equation for use within the model.

@@

\subsubsection{Cost Effectiveness}
\TODO{It also needs to talk about the cost effectiveness of reducing
  transmission or something, to say that we're getting competitive numbers. We
need a good source for cpca data for contact investigations, etc.}
We also estimated the cost per case averted of curing entering cases of LTBI. To
do so, we assumed that curing a single case of entering LTBI would cost \$800.
This number is substantially higher than the assumed domestic LTBI treatment
cost for a number of reasons; notably the cost of implementation as part of the
immigration process would be very nontrivial. However, this inflation is also
meant to be a conservative estimate, so as to underestimate the effectiveness of
this intervention rather than overestimate. Given this treatment cost, we found
that curing 25\% of entering cases of LTBI resulted in a net US HCS cost per
case averted of \$67,654.07 at 2025, \$28,699.15 at 2050, and
\$17,912.95 at 2100 (see Table \ref{tab:redEnLTBICosts} for the cost per case
averted for other cure percentages). Given the variable nature of LTBI treatment
cost, the model code is built so that a user can adjust these costs themselves
to explore more specific methods of curing entering LTBI, especially as the
treatment methods and associated costs evolve in future years. Curing
immigrating cases of LTBI  also resulted in 11,900, 29,880, and 60,189 fewer
cases of TB and 1,025, 2,573, and 5,185 fewer TB deaths for 10\%, 25\%, and 50\%
cure rate, respectively.

%We also found that the relationship between total incidence at 2100 and
%percentage of incoming LTBI cases cured was linear. Given this, we estimated the
%yearly average US HCS savings garnered by curing one case of entering LTBI per
%year from 2000 to 2025, 2000 to 2050, and 2000 to 2100.  This value peaked
%at \$1.283 billion at 2100 (25\% reduction). This illustrates that it
%would be cost saving to cure cases of LTBI at the cost of \$1.283 billion
%``2000'' dollars over the time period 2000-2100. 
%These findings are illustrated in Table~\ref{tab:redEnLTBICosts}

@@

@@

The extended Hill also added certain equations and constants to generate cost
data. These equations are detailed in Figure~%\ref{fig:hillCosts}. 
\TODO{Add figure hillCosts. We first need to determine what the correct LTBI
cost parameters are. See the TODO above regarding that.}
These parameters were derived via
\TODO{Cite Dylan's cost study, adherence study, etc.}

>>>>

The majority of TB infections are caused by Mycobacterium tuberculosis, 
an airborne pathogen which primarily affects the lungs and pulmonary system.  
Individuals infected with this bacteria first develop a latent form of TB
referred to as latent tuberculosis infection (LTBI).  These individuals are
asymptomatic and not contagious, and thus, this infection is difficult to detect
and can persist for many years. The latent tuberculosis infection may progress
to active tuberculosis infection (active TB), at which point the patient becomes
symptomatic and contagious. This progression is most likely within the first two
years after contraction, but can happen throughout the rest of the lifespan of
the individual.  Left untreated, individuals with LTBI have a lifelong
probability of approximately 10\% to develop active TB. The health risks for
individuals with active TB are severe, and without medical interventions, the 10
year mortality rate for the disease is between 20\% and 70\%, depending on the
severity of the case and definition of a positive diagnosis
\cite{_who_2013,tufariello_latent_2003}.

@@

The model was also extended to provide similarly sourced US health care system
(HCS) TB costs. To do so, we made the following treatment cost assumptions. A
single case of active TB was assumed to yield a \$14,000 US HCS cost, charged
immediately upon disease contraction. This cost is the weighted mean of the
costs of cases requiring hospitalization (49\%) and cases not requiring
hospitalization (51\%).  LTBI treatment costs were assumed to be \$468.00,
charged immediately upon successful treatment. These estimates were calculated
based on the cost of a successful treatment, and the typical adherence and
efficacy of said treatment
\cite{lobue_use_2003,page_kr_improved_2006,denholm_use_2010}.
\TODO{<+DYLAN COST STUDY CITATION+>.} 
Note
that active TB treatment costs are charged immediately upon disease contraction,
whereas LTBI treatment costs are charged only upon successful treatment. The
motivation for this distinction is that active TB treatment is mandated by the
US for all known cases and, further, most treatment costs are incurred at the
beginning of the treatment cycle. As such, charging for every known case,
immediately at contraction is a well motivated simplification. On the other
hand, latent TB is only rarely treated in discovered cases and the costs are
more evenly distributed. Furthermore, the choice to charge only upon successful
treatment yields a more conservative estimate of the success of any intervention
analyzed, as it will always underestimate the total US HCS TB cost burden.
Finally, it is worth noting that these estimates of US HCS treatment cost are
not intended to be fixed parameters. The code for this model is freely
available, and as new research amends the necessary costs and new methods of
treatment are developed, these values can be modified and the model re-run to
produce a revised prediction of the US HCS TB cost. In addition to US HCS cost,
the system was implemented so as to also track projected intervention
implementation cost, given user-inputted parameters relating to various possible
intervention cost strategies.  Cured cases of LTBI entering, TB deaths, total TB
cases, and total US HCS cost were also tracked discounted at 3\% annually for
cost benefit analysis. This discounting was converted to a continuous
differential equation for use within the model.

@@

\subsubsection{Cost Effectiveness}
We also estimated the cost per case averted of curing entering cases of LTBI. To
do so, we assumed that curing a single case of entering LTBI would cost \$800.
This number is substantially higher than the assumed domestic LTBI treatment
cost for a number of reasons; notably the cost of implementation as part of the
immigration process would be very nontrivial. However, this inflation is also
meant to be a conservative estimate, so as to underestimate the effectiveness of
this intervention rather than overestimate. Given this treatment cost, we found
that curing 25\% of entering cases of LTBI resulted in a net US HCS cost per
case averted of \$67,654.07 at 2025, \$28,699.15 at 2050, and
\$17,912.95 at 2100 (see Table \ref{tab:redEnLTBICosts} for the cost per case
averted for other cure percentages). Given the variable nature of LTBI treatment
cost, the model code is built so that a user can adjust these costs themselves
to explore more specific methods of curing entering LTBI, especially as the
treatment methods and associated costs evolve in future years. Curing
immigrating cases of LTBI  also resulted in 11,900, 29,880, and 60,189 fewer
cases of TB and 1,025, 2,573, and 5,185 fewer TB deaths for 10\%, 25\%, and 50\%
cure rate, respectively. These costs compare favorably to certain styles of
contact investigation, which is a widely practiced intervention. In particular,
using the common concentric circle approach to contact investigation over the contact
priority method (which is slightly less effective, but cheaper) has a cost per case
averted of \$92,934 from one estimate, which is substantially more than the cost
per case averted of treating entering LTBI found here \cite{pisu_targeted_2009}.

@@

\TODO{Improve Figure~\ref{fig:finalDists}. The labels need to be bigger, and it
might be wise to illustrate via some marker where the hill model results live.}

@@

The extended Hill also added certain equations and constants to generate cost
data. These equations are detailed in Figure~\ref{fig:hillCosts}.
\begin{figure}[h]
  \centering
  \TODO{Add this figure. We first need to determine what the correct LTBI
  cost parameters are. See the TODO above regarding that. Also cite Dylan's
paper here.}
  \caption{The equations used to track the economic impact of US TB dynamics.
  Parameters derived principally from
  \cite{lobue_use_2003,page_kr_improved_2006,denholm_use_2010}}
  \label{fig:hillCosts}
\end{figure}
